---
title: "Appendices"
output: pdf_document
---

## Appendix A 


Under the assumption of normality for all predictors (in each class), the concordance statistic ($C$) of the data can be calculated directly, using equation (3) [@mvauc].  This equation is suitable when the covariance matrices of each class are *not* equivalent. For $p$ predictors, $\pmb{\Delta_\mu}$ is a $p$ x $1$ vector housing the differences in predictor means between class 0 and class 1. $\pmb{\Sigma_0}$ and $\pmb{\Sigma_1}$ represent the covariance matrices of class 0 and 1 respectively and $\Phi$ is the cumulative density function (cdf) of the standard normal distribution.\

\begin{equation} \tag{3}
C = \Phi \left( \sqrt{\pmb{\Delta_\mu}{'}\  (\pmb{\Sigma_0} + \pmb{\Sigma_1})^{-1} \ \pmb{\Delta_{\mu}}} \right)
\end{equation}
\
\
In our research, the differences in predictor means between the classes were equivalent. In other words, the elements of $\pmb{\Delta_\mu}$ were equivalent; denoted by $\delta_\mu$.  The differences in predictor variances between the classes were also equivalent.  The diagonal elements of $\pmb{\Sigma_0}$ were all zero, and the diagonal elements of $\pmb{\Sigma_1}$ were all equal to ($1- \delta_\Sigma$) as shown in section 2.1.  To ensure a unique solution, $\delta_\Sigma$ was fixed to $0.3$.  Then, equation (1) was solved to determine the value of $\delta_\mu$ that yields a concordance statistic ($C$) of $0.85$. \
\
Let $\pmb{A} = (\pmb{\Sigma_0} + \pmb{\Sigma_1})^{-1}$,

\begin{align*}
(\Phi^{-1}(C))^2 &= \pmb{\Delta_\mu{'}}  \pmb{A} \  \pmb{\Delta_{\mu}}\\
(\Phi^{-1}(C))^2 &= 
\begin{bmatrix}
 \delta_\mu & \delta_\mu &   \dots &  \delta_\mu &  \delta_\mu
\end{bmatrix} \begin{bmatrix} 
    a_{11} & a_{12}  & \dots  & a_{1p}\\
    \vdots &         & \ddots & \\
    a_{p1} &  a_{p2} & \dots  & a_{pp} 
\end{bmatrix}\begin{bmatrix}
 \delta_\mu \\ \delta_\mu \\ \vdots \\ \delta_\mu \\ \delta_\mu
\end{bmatrix}\\
(\Phi^{-1}(C))^2 &= \delta_{\mu}^2 \sum_{j = 1}^{p} \sum_{i = 1}^{p} a_{ij}
\end{align*}
\
\
\
Based on a desired $C$ of $0.85$, 

\begin{equation} 
\delta_\mu = \frac{\Phi^{-1}(0.85)}{\sqrt{\sum_{j = 1}^{p} \sum_{i = 1}^{p} a_{ij}}}.
\end{equation}\
\

Equation (2) was used to derive the appropriate $\delta_\mu$ for each simulation scenario. 

## Appendix B 

Errors and warnings were carefully monitored throughout our simulation.  We used the tryCatch.W.E function from `simsalapar` to save all warning and error messages [@simsalapar].  These messages were stored in a data file for each scenario and, along with the raw simulation results, are available upon request.  In Table BX and Table BY we present the frequency with which errors occurred for the imbalance corrections and classification algorithms, respectively.\
\
Under scenarios where training data were class balanced ($EF = 0.5$), imbalance correction methodologies did not preform reliably.  For instance, if the classes were exactly balanced there is nothing to over or under sample, and the algorithm returned an error. In the event of an error with imbalance corrections, the uncorrected data were allowed to proceed to train the classification algorithms.\
\
Conversely, when data were extremely imbalanced ($EF = 0.02$) and sample size was small ($\frac{1}{2}N$ some classification algorithms returned an error. In this case, predicted risks were stored as `NA`. In case of data separation, warnings messages were stored and predicted risks were saved as usual. Finally, under extreme imbalance, the algorithm RUSBoost produced impossible predicted risks (i.e., outside the range $[0,1]$), in these instances, predicted risks below zero and above one were set to $1e-10$ and $1 - e-10$, respectively.\
\
```{r, echo = F}
# corrections
cr <- readRDS("error_table_corrections.RDS")

# algorithms
lr <- readRDS("error_table_logistic_regression.RDS")
sv <- readRDS("error_table_support_vector_machine.RDS")
rf <- readRDS("error_table_random_forest.RDS")
xg <- readRDS("error_table_xgboost.RDS")
rb <- readRDS("error_table_rusboost.RDS")
ee <- readRDS("error_table_easy_ensemble.RDS")
```

```{r, echo = F}
library(xtable)
library(kableExtra)

cr %>% kable() %>% kable_styling()
lr %>% kable() %>% kable_styling()
sv %>% kable() %>% kable_styling()
rf %>% kable() %>% kable_styling()
xg %>% kable() %>% kable_styling()
rb %>% kable() %>% kable_styling()
ee %>% kable() %>% kable_styling()


# need to style the kables 
# and filter for non-zero entries :) 
```














