---
title: "Master's Thesis"
output: 
  pdf_document:
    citation_package: natbib

tables: true
cls: WileyNJD-v2.cls
bibliography: "bibliography.bib"
biblio-style: unsrt
link-citations: true
header-includes: 
   \usepackage[labelfont=bf]{caption}
   \usepackage[super, compress, comma]{natbib}
---

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(xtable)
```


# 1 | Introduction 

\
\


# 2 | Methods 

In this research, we demonstrated the effects of common class imbalance correction methodologies, when used in the context of clinical prediction model development.  We focused on clinical prediction models designed for dichotomous risk prediction.  A simulation study was implemented to illustrate the effects of imbalance correction methodologies across 18 unique data-generating scenarios (section 2.2).  For each scenario, we compared the out-of-sample predictive performance (i.e., model performance on data not used to train the model) of prediction models developed using a two-step procedure (section 2.3).  

## 2.1 | Data Generating Mechanism 

In this simulation study, we focused on prediction models for dichotomous risk prediction. Consequently, we generated data comprised of two classes.  We refer to the negative class (non-events) and positive class (events) as class 0 and class 1, respectively.  Data for each class were generated independently using distinct multivariate normal ($mvn$) distributions.  As shown in equations 1 and 2, we specified a distinct mean and covariance structure for each class. The mean structures for the classes are represented as $\mu_0$ and $\mu_1$ for class 0 and 1, respectively. The covariance matrices are represented as $\Sigma_0$ and $\Sigma_1$ for class 0 and 1, respectively.

\begin{align}
&\mathrm{Class \ \  0:} \mathbf{X} \sim mvn( \pmb{\mu_0}, \pmb{\Sigma_0}) = mvn(\pmb{0}, \pmb{\Sigma_0}) \\
&\mathrm{Class \ \  1:} \mathbf{X} \sim mvn( \pmb{\mu_1}, \pmb{\Sigma_1}) = mvn(\pmb{\Delta_\mu}, \pmb{\Sigma_0} - \pmb{\Delta_\Sigma})
\end{align}


As shown in equation 2, the differences in parameter values between the two classes are represented by $\pmb{\Delta_\mu}$ and $\pmb{\Delta_\Sigma}$; a vector and matrix comprised of the differences in predictor means, and variances/ covariances, between the classes, respectively.  We specified no variation in means among predictors within a class, making all elements in $\pmb{\Delta_\mu}$ equivalent; we denote these equivalent elements as $\delta_\mu$. Similarly, we specified no variation in predictor variances within a class, making all diagonal elements in the matrix $\pmb{\Delta_\Sigma}$ equivalent; diagonal elements are denoted by $\delta_\Sigma$.\
\
For class 0, all predictor means were fixed to zero and all variances to 1. For class 1, all means were non-zero and are represented by the vector $\pmb{\Delta_\mu}$ (since class 0 means are all zero). Finally, in both classes, we allowed $80$% of the predictors to covary. All non-zero correlations among predictors in each class were set to $0.2$.  To ensure the correlation among predictors was not stronger in one class, we fixed the correlation matrices of the two classes to be equal.  This was accomplished by computing the off-diagonal elements of $\pmb{\Delta_\Sigma}$ such that the correlation matrices of the two classes were equivalent (as shown below)\

For instance, with 8 predictors:\
\
the mean and covariance structure for class 0 is, \begin{equation*}
\pmb{\mu_0} = \begin{bmatrix}
 \\ 0 \\ 0 \\ 0\\ 0 \\ 0 \\ 0 \\ 0 \\ 0
\end{bmatrix}, \pmb{\Sigma_0} = \begin{bmatrix}
1   & 0.2 & 0.2 & 0.2 & 0.2 & 0.2 & 0 & 0\\
0.2 & 1   & 0.2 & 0.2 & 0.2 & 0.2 & 0 & 0\\
0.2 & 0.2 & 1   & 0.2 & 0.2 & 0.2 & 0 & 0\\
0.2 & 0.2 & 0.2   & 1 & 0.2 & 0.2 & 0 & 0\\
0.2 & 0.2 & 0.2   & 0.2 & 1 & 0.2 & 0 & 0\\
0.2 & 0.2 & 0.2   & 0.2 & 0.2 & 1 & 0 & 0\\
0   & 0   & 0     &  0 & 0    & 0 & 1 & 0\\
0   & 0   & 0     &  0 & 0    & 0 & 0 & 1\\
\end{bmatrix}
\end{equation*}

and mean and covariance structure for class 1 is, \begin{equation*}
\pmb{\mu_1} = \begin{bmatrix}
 \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu \\ \delta_\mu
\end{bmatrix}, \pmb{\Sigma_1} = \begin{bmatrix}
1 - \delta_\Sigma   & z & z & z & z & z & 0 & 0\\
z & 1 - \delta_\Sigma   & z & z & z & z & 0 & 0\\
z & z & 1 - \delta_\Sigma   & z & z & z & 0 & 0\\
z & z & z   & 1 - \delta_\Sigma & z & z & 0 & 0\\
z & z & z   & z & 1 - \delta_\Sigma & z & 0 & 0\\
z & z & z   & z & z & 1 - \delta_\Sigma & 0 & 0\\
0   & 0   & 0     &  0 & 0    & 0 & 1 - \delta_\Sigma & 0\\
0   & 0   & 0     &  0 & 0    & 0 & 0 & 1 - \delta_\Sigma\\
\end{bmatrix}.
\end{equation*}\
\
Here, $z = (1-\delta_\Sigma)*0.2$, to ensure equivalent correlation matrices between the two classes.  Note, the covariance matrices were *not* equivalent between the classes.\
\
In this simulation study, 18 unique data-generating scenarios were selected (see Section 2.2).  Parameter values for the data generating distributions ($\delta_\mu$ and $\delta_\Sigma$) of each scenario were selected to generate a concordance statistic ($C$) of $0.85$. Under the assumption of normality for all predictors (in each class), the concordance statistic of the data can be expressed as a function of $\pmb{\Delta_\mu}$, $\pmb{\Sigma_0}$ and $\pmb{\Sigma_1}$ [@mvauc]. Optimal values of $\delta_\mu$ and $\delta_\Sigma$ for each scenario were computed analytically, based on the following formula [@mvauc]:

\begin{equation}
C = \Phi \left( \sqrt{\pmb{\Delta_\mu}{'}\  (\pmb{\Sigma_0} + \pmb{\Sigma_1})^{-1} \ \pmb{\Delta_{\mu}}} \right).
\end{equation}

In equation (3), $\Phi$ represents the cumulative density function of the standard normal distribution; $\pmb{\Delta_\mu}$, $\pmb{\Sigma_0}$ and $\pmb{\Sigma_1}$ maintain their previous definitions. To ensure a unique solution, $\delta_\Sigma$ was fixed at 0.3 for each scenario, while equation (3) was solved to yield the appropriate value of $\delta_\mu$ in each scenario. Full analytical solution is provided in Appendix A.\
\
Finally, given that data for each class were generated independently, we had direct control over how many observations were generated under each class. The number of observations from the positive class ($n_1$) was sampled from the binomial distribution with probability equal to the specified event fraction. The number of observations in the negative class ($n_0$) was then computed as $N - n_1$, where $N$ is the specified sample size.\
\
Parameter values for the data generating distributions of the simulation scenarios are presented in Table 1.

## 2.2 | Data Generating Scenarios 

In our simulation study, 18 ($3$ x $2$ x $2$) unique data-generating scenarios were implemented (Table 1). This was achieved by varying the following three characteristics of the data: the event fraction (proportion of patients with an event), the number of predictors and sample size. Event fraction (EF) varied through the set {0.5, 0.2, 0.02} and number of predictors through the set {8, 16}. In all scenarios, data were generated to yield an expected concordance statistic ($C$) of $0.85$.  Given the number of predictors, the event fraction and expected concordance statistic, we computed the minimum required sample size for a prediction model developed under these conditions.  Sample size calculations were carried out using the R package pmsampsize [@pmsampsize].  The sample size of data used to train the prediction models was then varied through the set {$\frac{1}{2}N$, $N$, $2N$} implying half the required sample size, adequate samples size and double the required sample size, respectively.\
\

```{r, echo = F, results = "asis"}
# .csv containing data frame of simulation scenarios
sim_sets <- read.csv("./../data_generating_mechanism/sim_settings.csv") %>% 
  filter(npred < 17)

# .Rdata containing data frame of data generating parameters and sample size
#  for each scenario
set <- readRDS("./../data_generating_mechanism/set.RData")

# retrieve sample size for each scenario
sample_size <- c()

for (i in 1:18){
  sample_size[i] <- set[[i]][[4]]
}

sim_sets <- 
  sim_sets %>% 
  as.data.frame() %>% 
  mutate(
    sample_size = sample_size, 
    validation  = sample_size*10,
    sample_size = as.character(sample_size), 
    validation  = as.character(validation),
    auc = c(rep(0.85, 18)),
    n   = recode(n, 
                 "0.5" = "0.5N", 
                  "1"   = "N", 
                  "2"   = "2N")
    ) %>%
  dplyr::select(X, npred, n, ef, sample_size, validation, dmu, auc) %>%
  rename(
    "Scenario"         = X,
    "Event Fraction"   = ef, 
    "No. Predictors"   = npred,
    "Sample Size"      = n,
    "$\\mathrm{n}_{\\mathrm{train}}$"      = sample_size,
    "$\\mathrm{n}_{\\mathrm{validation}}$" = validation, 
    "$\\delta_\\mu$"   = dmu,
    "C"                = auc) 

sim_sets %>% 
  xtable(escape = F,
         caption = "Summary of simulation scenarios.", label = "tab:table") %>% 
  print(comment = F,
        booktabs = T,
        sanitize.text.function = function(x) {x},
        include.rownames=FALSE)
```

\

## 2.3 | Model Development 

All prediction model in our research were developed according to this  two-step procedure.  First, data are pre-processed using an imbalance correction technique. Then, the resulting pseudo-balanced data are  used to train a classification algorithm.\
\
We implemented a $5$ x $6$ factorial design, to compare the out-of-sample predictive performance of prediction models developed with $5$ imbalance corrections ($1$ control and $4$ imbalance corrections) and $6$ classification algorithms.  In total, we compare the performance of 30 prediction models, each comprised of a unique combination of an imbalance correction and classification algorithm.

\newpage

### 2.3.1 | Imbalance Corrections 

The imbalance corrections studied in our simulation included: control, random under sampling (RUS), random over sampling (ROS), synthetic majority over sampling (SMOTE) and synthetic majority over sampling with edited neared neighbors (SENN).  **more justification for inclusion** . As determined by a recent systematic review, the most common class imbalance correction currently implemented when developing clinical prediction models is SMOTE [@constanza].  All imbalance corrections and the R packages used for their implementation are summarized in Table 2.\
\
In the control "correction", data were not corrected for imbalance, and moved to second step of model development untouched (the imbalanced data were used to train prediction models).  All imbalance correction methodologies were implemented to achieve artificially class balanced data ($EF_\mathrm{corrected} \approx 0.5$). RUS achieves class balanced data by randomly disregarding observations from the majority class such that the resulting data are class balanced.  ROS generates artificial observations for the minority class by re-sampling observations from the minority class with replacement until a balance is achieved.  SMOTE generates artificial observations from the minority class by ---- [@citation]; in our implementation we specify the number of nearest neighbors for smote (k) to be 5, the default in the package used for implementation [@iric]. SMOTE - ENN --- works like this [@citation].  We specify the number of nearest neighbors in the SMOTE step (k1) to be 5, and the number of nearest neighbors in ENN steep (k2) to be 3, the software defaults [@iric]. 

### 2.3.2 | Classification Algorithms 

Classification algorithms were selected based on a systematic review identifying common algorithms used to develop prediction models in a medical context [@constanza]. These algorithms include: logistic regression, support vector machine, random forest and XGBoost.  Additionally, based on literature summarizing common strategies to handle class imbalance [@yu; @summary_m; @lp; @kaur], we included two ensemble learning algorithms designed specifically to handle class imbalance: RUSBoost and EasyEnsemble. The hyperparameters selected for these classification algorithms, and the R packages used for algorithm implementation are summarized in Table 2.\

**Paragraph about implementation.** 
* lr - no hyperparameter 
* svm - caret, svmRadial, caret default tune grid (C), tunelength = 3 (default)
* rf - caret,custom tune grid - mtry (1:npred), min.node.size (min.node.size = 1:10), splitrule = "gini"
* xg - caret default tunegrid, tunelength = 3 (default)
* rusb - weak classifier = svm with radial kernel
* ee - nothing to tune here 


```{r, echo = F, results = "asis", warning = F, message = F}

Summary <- c("\\bf{Imbalance} \\\\ 
              \\bf{Corrections:}",
              " ", " ", " ", " ",
             "\\bf{Classification} \\\\ 
              \\bf{Algorithms:}",
               " ", " ", " ", " ", " ")

Method <- c(
  "Random Undesampling", 
  "Random Oversampling",
  "Synthetic Majority Over Sampling",
  "SMOTE - Edited Nearest Neighbours",
  " ", 
  "Logistic Regression",
  "Support Vector Machine",
  "Random Forest",
  "XGBoost",
  "RUSBoost",
  "EasyEnsemmble")

Abbreviation <- c("RUS","ROS", "SMOTE", "SENN"," ", "LR", "SVM", "RF", "XG", "RB", "EE")

Hyperparameters <- c( " ", " ", "$k=5$", "$k1=5$, $k2 = 3$", 
                     " ", " ", "kernel = radial, tune grid = A", "tune grid = B", "tune grid = C", "classifier = svm (radial)", " ") 

Implementation <- c( 
  "ROSE \\cite{rose}", 
  "ROSE \\cite{rose}", 
  "IRIC \\cite{iric}", 
  "IRIC \\cite{iric}",
  " ", 
  "base R \\cite{r}", 
  "caret \\cite{caret}", 
  "caret \\cite{caret}", 
  "caret \\cite{caret}",
  "ebmc \\cite{ebmc}",
  "IRIC \\cite{iric}")

cbind(Summary, Method, Abbreviation, Hyperparameters, Implementation)%>%
  as.data.frame()%>%
  rename("R Package" = "Implementation", 
         " " = "Summary")%>% 
  xtable(caption = "Summary of imbalance corrections and classification algorithms implemented.", 
         label = "tab:table", escape = F)%>% 
  print(comment = F,
        booktabs = T,
        sanitize.text.function = function(x) {x},
        include.rownames=FALSE)

  # kable(booktabs = T, escape = F,
  #       linesep = "\\addlinespace",
  #       caption = "Summary of classification algorithms used in simulation study.") %>%
  # kable_styling(full_width = F, latex_options ="hold_position", font_size = 10) %>% 
  # column_spec(1, width = "20em") %>%
  # column_spec(2, width = "7em") %>%
  # add_indent(c(1:10), level_of_indent = 2) %>%
  # pack_rows("Imbalance Corrections", 1, 4) %>%
  # pack_rows("Classification Algorithms", 5, 10)
```

## 2.4 | Simulation Methods

Under each simulation scenario, $2000$ data sets were generated. Each data set was comprised of training and validation data. The training and validation data were generated independently using identical data generating mechanisms. Validation data sets were generated to be ten times larger than the training data sets. The sample sizes of the training and validation data for each simulation scenario can be found in Table 1.\
\
For each generated data set, 30 ($5$ x $6$) prediction models were developed (as described in section 2.3).  All prediction models were trained using the training data. Out-of-sample performance was then assessed using the validation data.\
\
This simulation study was conducted using High Performance Computing -- ** insert tech specs**. 


## 2.4 | Performance Metrics 

Out-of-sample model performance was assessed using measures of calibration, discrimination and overall performance. All performance metrics were computed using validation data.\
\
Calibration was measured using visual and empirical metrics.  We assessed calibration visually by means of flexible calibration plots.  For each simulation iteration, a flexible calibration curve was generated with loess regression using ggplot2, for all prediction models [@gg].  Additionally, calibration intercept and slope were calculated according to their respective definitions in Steyerberg et al. (2010) [@epi].\
\
The concordance statistic was used to measure model discrimination; computed using the R package pROC [@pROC]. For dichotomous risk prediction, it is equivalent to the area under the Receiver Operator Characteristic curve [@epi].\
\
Overall performance was measured by scaled Brier score. This metric reflects both model discrimination and calibration and was calculated according its definition in Steyerberg et al. (2010) [@epi].\
\
For empirical measures of model performance (concordance statistic, scaled Brier score, calibration intercept and calibration slope), the median over the simulation iterations and corresponding Monte Carlo error were reported.

## 2.5 | Error Handling

Under scenarios where training data were class balanced ($EF = 0.5$), imbalance correction methodologies may not preform reliably.  For instance, if the classes are exactly balanced there is nothing to over or under sample, and the algorithm returns an error. In the event of an error with imbalance corrections, the uncorrected data were allowed to proceed to train the classification algorithms.  Conversely, when data were extremely imbalanced ($EF = 0.02$) and sample size is small ($\frac{1}{2}N$ some classification algorithms may return an error. In this case, predicted probabilties were stored as "$NA$".\
\
In our simulation, we save all warnings and errors, and report the frequency with which errors occur for each imbalance correction and algorithm in all simulation scenario. We used the tryCatch.W.E function from simsalapar to save errors and warnings in our simulation [@simsalapar]. 

* need to talk about recalibration somewhere * 


# 3 | Results 

# 4 | Disucssion

\newpage
\footnotesize