---
title: "errors"
output: html_document
---

#  1. rus and ros, throwing an error: 

- class balanced scenario, smallest sample size 
- not happening when there is class imbalance
```{r}
# Error in sample.int(length(x), size, replace, prob) : 
# cannot take a sample larger than the population when 'replace = FALSE'
```

```{r}
# SOLUTION (temporary):

# repeat rus or ros until it works 
```

* need to store indication of how often this happens ?
* need to store the error messages


```{r}
# demonstrate error 

do_rus <- function(dataframe){
  
  formula <- outcome~.
  data    <- dataframe
    
  df.rus <- ROSE::ovun.sample(formula, data, method = "under")$data
  
  return(df.rus)
}
```

```{r}
all_data <- readRDS("generated_data/scenario_1.rds")

data <- all_data$data

do_rus(data[[2]])
```


# 2. glm not converging when: 

- extreme class imbalance, smallest sample size - happens all the time 
- not happening when there is class balance

* due to separability of the classes
```{r}
# Warning: glm.fit: algorithm did not convergeWarning: glm.fit: fitted probabilities numerically 0 or 1 occurred
```

```{r}
# SOLUTION: 

# use TryCatch to detect warning 
# Then save as indicator 
```

* need to store the warnings


# 3. Need to order factor -- or sometime R sees 0 as outcome to model 

```{r}
data[[2]]$outcome = factor(data[[2]]$outcome, levels=c(0,1), ordered = TRUE) 
```

# 4. For sc 3, pair 8, 

all of sudden this is happening 
```{r}
# Warning: There were missing values in resampled performance measures.Warning: model fit failed for Fold1.Rep1: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold3.Rep1: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold1.Rep2: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold2.Rep2: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold3.Rep2: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold4.Rep3: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: model fit failed for Fold5.Rep3: C=1 Error in .local(object, ...) : test vector does not match model !
# Warning: There were missing values in resampled performance measures.
```

-- investigate this. 


# 5. Smote from smote family has problems -- 







