---
title: "Data Generating Mechanism"
author: "Alex Carriero"
output: html_document
---

Please find our data-generating mechanism detailed below: 

# 1. Set Up
```{r, warning = F, message = F}

# load libraries
library(tidyverse)
library(MASS)
library(matlib)
library(tableone)
library(pmsampsize)
library(pROC)

```

# 2. Define functions necessary for data-generating function 
```{r}
sample_size <- function(npred, ev, n.level){
  
  # this function calculates and returns the minimum required sample size (N) for a prediction model 
  #
  # npred   : number of predictors 
  # ev      : event fraction 
  # n.level : the value the minimum sample size should be multiplied by
  #           to achieve the desired sample size setting (i.e., 0.5N, N, 2N)
  
  a = pmsampsize(type = "b", prevalence = ev, cstatistic = 0.85, parameters=npred)$sample_size
  n = ceiling(a*n.level)

  set.seed(NULL)
  return(n)
  
}
```

```{r}
scenario <- function(npred, ev = 0.5 , n.level = 1, dmu = 0.5){
  
  # this function takes each factor to be varied in simulation as input
  # and generates the mean and covariance matrices for each class.
  #
  # npred   : number of predictors 
  # ev      : event fraction 
  # n.level : the value the minimum sample size should be multiplied by
  #           to achieve the desired sample size scenario (i.e., 0.5N, N, 2N)
  #
  # it returns a list of all information required data generation in a given scenario: 
  # number of predictors, event fraction, n.level, sample size, 
  # and the mean and covariance matrices of each class. 

  # set up
  x      <- dmu                              # delta mu
  y      <- 0.3                              # delta sigma 
  z      <- 0.2                              # z 
  p      <- 0.75                             # proportion of correlated predictors
  npred  <- npred                            # number of predictors
  n      <- sample_size(npred, ev, n.level)  # sample size calculation

  # set up correlations
  corr0  <-  matrix(0, npred, npred)         # matrix: set up for cov matrix, 0 on diagonals
  corr0[1:npred*p, 1:npred*p] = z            # class 0 
  diag(corr0) = 0

  corr1  <-  matrix(0, npred, npred)         # matrix: set up for cov matrix, 0 on diagonals
  corr1[1:npred*p, 1:npred*p] = (1-y)*z      # class 1
  diag(corr1) = 0

  # mean structures
  mu0    <-  c(rep(0,npred))                 # vector: class 0 means
  dmu    <-  c(rep(x,npred))                 # vector: difference in means between classes 
  mu1    <-  mu0 + dmu                       # vector: class 1 means 

  # covariance structures
  sigma0 <-  diag(npred)  + corr0            # matrix: cov matrix of class 0 
  dsig   <-  diag(c(rep(y, npred)))          # matrix: difference in variances between classes 
  sigma1 <-  diag(npred) - dsig + corr1      # matrix: cov matrix of class 1

  return(list("npred"       = npred, 
              "event_frac"  = ev, 
              "n_level"     = n.level, 
              "sample_size" = n, 
              "mu0"         = mu0, 
              "mu1"         = mu1, 
              "sigma0"      = sigma0, 
              "sigma1"      = sigma1))
  
}
```

# 3. The data-generating function
```{r}
generate_data <- function(inn, test = FALSE){
  
  # this functions takes a list of data-generating parameters as input
  # and outputs one data set with class data.frame
  # 
  # when test = FALSE the function generates training data 
  # 
  # when test = TRUE the function generates validation data
  #
  # the data generating mechanism for training vs. test data is identical
  # these data sets only differ in sample size: 
  # validation data has 10x the sample size of training data
  
  
  npred   = inn[[1]]  # number of predictors
  ev      = inn[[2]]  # event fraction
  n.level = inn[[3]]  # 0.5, 1, 2 (to adjust sample size)
  n       = inn[[4]]  # minimum required sample size for the prediction model (N)
  mu0     = inn[[5]]  # mean structure for class 0 as vector
  mu1     = inn[[6]]  # mean structure for class 1 as vector
  sigma0  = inn[[7]]  # covariance structure for class 0 as matrix
  sigma1  = inn[[8]]  # covariance structure for class 1 as matrix
  
  # test set
  if (test == TRUE){n  = n*10}
  
  # positive class
  n1      <- rbinom(1, n, ev)
  class_1 <- mvrnorm(n1, mu1, sigma1)
  
  # negative class
  n0      <- n - n1
  class_0 <- mvrnorm(n0, mu0, sigma0)
  
  outcome <- c(rep(1, n1), rep(0, n0))
  
  # format data frame
  df <- cbind(outcome, rbind(class_1, class_0)) %>% 
    as.data.frame() %>% 
    mutate(outcome = as.factor(outcome))
  
  # names
  colnames(df) <- c("outcome", paste0("x", 1:npred))
  
  return(df)
}

```


# 4. A test example -- interact with simulated data!
```{r, warning = F, message = F}
# function to return summary statistics for each class 

summ <- function(df){
  a <- CreateTableOne(data = df, strat = "outcome")
  return(a)
}
```

```{r, warning = F, message = F}
# function to return gg scatter plot displaying data portrayed by 2-dimensions

visualize <- function(df){
  df <- df 
  df %>%
  arrange(outcome)%>%
  ggplot(aes(x = x1, y = x2, col = outcome)) + 
  geom_point(alpha = 0.7) + 
  theme_minimal() + 
  ggtitle("Simulated Data:") + 
  scale_color_manual("Class", values = c("goldenrod1", "darkblue")) + 
  xlab("Predictor 1") +
  ylab("Predictor 2")
}
```

```{r}
# check discrimination (concordance statistic calculated using logistic regression)

get_auc <- function(df){
  
mod  <- glm(outcome~., family = "binomial", data = df)
pred <- predict(mod, type="response")
auc  <- roc(df$outcome, pred, quiet = TRUE)$auc

return(auc)
}
```

## Test Example 
```{r}
# generate data as example
df  <- generate_data(scenario(npred = 8, ev = 0.2, n.level = 1, 0.604))
df
```

```{r}
# view data 
summ(df) 
get_auc(df)
visualize(df)
```

# 5. Get Parameter Values Necessary to Produce Desired AUC = 0.85
```{r}
get_delta_mu <- function(npred, auc = 0.85){
  # based equation A2 in Appendix A 
  
  inn = scenario(npred)      # get covariance matrices
  s0  = inn[[7]]              
  s1  = inn[[8]]                       
  
  A   = inv(s0 + s1)         # calculate matrix A
  Y   = sum(colSums(A))                  
  
  dmu = qnorm(auc) / sqrt(Y)
  return(dmu)
}
```

```{r}
get_delta_mu(npred = 8)   # 0.6043
get_delta_mu(npred = 16)  # 0.4854
```

# 6. Store simulation scenario settings: generate file sim_settings.csv
```{r}
# store scenario settings in a data frame
# i.e., create a data frame with 18 rows and 4 columns
#
# rows: one simulation scenario
# cols: number of predictors (8, 16), 
#       event fraction (0.5, 0.2, 0.02)
#       sample size (0.5N, N, 2N)
#       dmu (0.6042, 0.4854)
#       where, dmu represent delta_mu
#       ... the mean structure of class 1 that result in a concordance statistic of 0.85

ef  <- c(0.5, 0.2, 0.02)
p   <- c(8, 16)
n   <- c(0.5, 1, 2)

grid <- expand.grid(ef, n, p)
grid <- 
  grid %>% 
  mutate(dmu = c(rep(0.6043, 9), rep(0.4854, 9))) %>% 
  rename("ef"      = Var1, 
         "n"       = Var2, 
         "npred"   = Var3)%>% 
  dplyr::select(npred, ef, n, dmu)

# ... and store in a matrix 
gridmat <- data.matrix(grid)
```

```{r}
# view scenario settings
grid
```

```{r}
# save settings in csv
write.csv(grid,"sim_settings.csv")
```

# 7. Store data-generating parameters for each simulation scenario: generate file set.RData
```{r}

# This code creates a list with 18 objects .. one for each simulation scenario 
#
# each of the 18 objects represent the output of the function scenario()
# defined in step 2. 

set <- list()  

for(i in 1:nrow(gridmat)){
  colnames(gridmat)<-NULL
  set[[i]] <- scenario(npred   = gridmat[i,1], 
                       ev      = gridmat[i,2], 
                       n.level = gridmat[i,3], 
                       dmu     = gridmat[i,4])
}

saveRDS(set, file="set.RData")
```

```{r}
# view data generating parameters for one scenario 
set[[18]]
```

# 8. Checks

```{r, message = F}
# a small simulation to see mean and sd in concordance statistic for each scenario -- using logistic regression
# to test that on average the concordance statistic is 0.85

set.seed(1111)

mean_auc <- c()
sd_auc   <- c()
  
for(j in 1:length(set)){
  scenario   <- set[[j]]
  auc        <- c()

  for (i in 1:2000){    
    df      <- generate_data(scenario)
    auc[i]  <- get_auc(df)
  }

mean_auc[j] <- mean(auc)   
sd_auc[j]   <- sd(auc)
}
```

```{r}
# check that correlation matrix is the same for class 1 and class 0 
a<- set[[2]][[7]] # class 0
b<- set[[2]][[8]] # class 1 

a
b
cov2cor(a)
cov2cor(b)
# yes 
```




