---
title: "extras from sim"
output: html_document
---

# The Article Header Information

YAML header:

```
output:
  rticles::sim_article:
    keep_tex: TRUE
```

Configure the YAML header including the following elements:

* `title`: Title

* `author`: List of author(s) containing `name` and `num`

* `address`: List containing `num` and `org` for defining `author` affiliations

* `presentaddress`: Not sure what they mean with this

* `corres`: Author and address for correspondence

* `authormark`: Short author list for header

* `received`, `revised`, `accepted`: dates of submission, revision, and acceptance of the manuscript

* `abstract`: Limited to 250 words

* `keywords`: Up to 6 keywords

* `bibliography`: BibTeX `.bib` file

* `classoption`: options of the `WileyNJD-v2` class

* `longtable`: set to `true` to include the `longtable` package, used by default from `pandoc` to convert markdown to \LaTeX code

## Remarks

1. In `authormark` use _et al._ if there are three or more authors.

2. Note the use of `num` to link names and addresses.

3. For submitting a double-spaced manuscript, add `doublespace` as an option to a `classoption` line in the YAML header: `classoption: doublespace`.

4. Keywords are separated by semicolons.

# The Body of the Article

## Mathematics

Use mathematics in Rmarkdown as usual.

## Figures and Tables

Figures are supported from R code:

```{r plot-ref, fig.cap = "Fancy Caption\\label{fig:plot}"}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
```

...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`. It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323).

Analogously, use Rmarkdown to produce tables as usual:

```{r, results = "asis"}
if (!require("xtable")) install.packages("xtable")
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
```

Referenced via \ref{tab:table}. You can also use the YAML option `header-includes` to includes custom \LaTeX packages for tables (keep in mind that `pandoc` uses `longtables` by default, and it is hardcoded; some things may require including the package `longtable`). E.g., using `ctable`:
```
header-includes:
- \usepackage{ctable}
```
Then, just write straight-up \LaTeX code and reference is as usual (`\ref{tab:ctable}`):
```
\ctable[cap = {Short caption},
        caption = {A long, long, long, long, long caption for this table.},
        label={tab:ctable},]
        {cc}
        {
        \tnote[$\ast$]{Footnote 1}
        \tnote[$\dagger$]{Other footnote}
        \tnote[b]{Mistakes are possible.}
        }{
        \FL
        COL 1\tmark[a] & COL 2\tmark[$\ast$]
        \ML
        6.92\tmark[$\dagger$] & 0.09781 \\
        6.93\tmark[$\dagger$] & 0.09901 \\
        97 & 2000
        \LL
}
```

It is also possible to set the `YAML` option `longtable: true` and use markdown tables (or the `knitr::kable` function): `knitr::kable(head(cars))` produces the same table as the `xtable` example presented before.

## Cross-referencing

The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
for figures, tables, equations, etc., is encouraged (using `[@<name>]`, equivalent of `\ref{<name>}` and `\label{<name>}`). That works well for citations in Rmarkdown, not so well for figures and tables. In that case, it is possible to revert to standard \LaTeX syntax.

## Double Spacing

If you need to double space your document for submission please use the `doublespace` option in the header.

# Bibliography

Link a `.bib` document via the YAML header, and bibliography will be printed at the very end (as usual). The default bibliography style is provided by Wiley as in `WileyNJD-AMA.bst`, do not delete that file.

Use the Rmarkdown equivalent of the \LaTeX citation system using `[@<name>]`. Example: [@Taylor1937], [@Knupp1999; @Kamm2000].

To include all citation from the `.bib` file, add `\nocite{*}` before the end of the document.

# Further information

All \LaTeX enviroments supported by the main template are supported here as well; see the `.tex` sample file [here](http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1097-0258/homepage/la_tex_class_file.htm) for more details and example.


\begin{table}[ht]
\centering
\begin{tabular}{rrlrllrrr}
  \toprule
Scenario & No. Predictors & Sample Size & Event Fraction & $\mathrm{n}_{\mathrm{train}}$ & $\mathrm{n}_{\mathrm{validation}}$ & $\delta_\mu$ & $\delta_\Sigma$ & C \\ 
  \midrule
  1 &   8 & 0.5N & 0.50 & 193 & 1930 & 0.60 & 0.30 & 0.85 \\ 
    2 &   8 & 0.5N & 0.20 & 124 & 1240 & 0.60 & 0.30 & 0.85 \\ 
    3 &   8 & 0.5N & 0.02 & 899 & 8990 & 0.60 & 0.30 & 0.85 \\ 
    4 &   8 & N & 0.50 & 385 & 3850 & 0.60 & 0.30 & 0.85 \\ 
    5 &   8 & N & 0.20 & 247 & 2470 & 0.60 & 0.30 & 0.85 \\ 
    6 &   8 & N & 0.02 & 1797 & 17970 & 0.60 & 0.30 & 0.85 \\ 
    7 &   8 & 2N & 0.50 & 770 & 7700 & 0.60 & 0.30 & 0.85 \\ 
    8 &   8 & 2N & 0.20 & 494 & 4940 & 0.60 & 0.30 & 0.85 \\ 
    9 &   8 & 2N & 0.02 & 3594 & 35940 & 0.60 & 0.30 & 0.85 \\ 
   10 &  16 & 0.5N & 0.50 & 193 & 1930 & 0.49 & 0.30 & 0.85 \\ 
   11 &  16 & 0.5N & 0.20 & 247 & 2470 & 0.49 & 0.30 & 0.85 \\ 
   12 &  16 & 0.5N & 0.02 & 1797 & 17970 & 0.49 & 0.30 & 0.85 \\ 
   13 &  16 & N & 0.50 & 385 & 3850 & 0.49 & 0.30 & 0.85 \\ 
   14 &  16 & N & 0.20 & 493 & 4930 & 0.49 & 0.30 & 0.85 \\ 
   15 &  16 & N & 0.02 & 3593 & 35930 & 0.49 & 0.30 & 0.85 \\ 
   16 &  16 & 2N & 0.50 & 770 & 7700 & 0.49 & 0.30 & 0.85 \\ 
   17 &  16 & 2N & 0.20 & 986 & 9860 & 0.49 & 0.30 & 0.85 \\ 
   18 &  16 & 2N & 0.02 & 7186 & 71860 & 0.49 & 0.30 & 0.85 \\ 
   \bottomrule
\multicolumn{4}{l}{\footnotesize Note: $N$ is the minimum required sample size for the prediction model.} \\
\end{tabular}
\caption{Summary of simulation scenarios.} 
\label{tab:sim_sets}
\end{table}


\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \toprule
Method & Abbreviation & Hyperparameters & R Package \\ 
  \midrule
Logistic Regression & LR &   & base R \cite{r} \\ 
  Support Vector Machine & SVM & tune grid = A & caret \cite{caret} \\ 
  Random Forest & RF & tune grid = B & caret \cite{caret} \\ 
  XGBoost & XG & tune grid = C & caret \cite{caret} \\ 
  RUSBoost & RB &   & ebmc \cite{ebmc} \\ 
  EasyEnsemmble & EE &   & IRIC \cite{iric} \\ 
   \bottomrule
\end{tabular}
\caption{Summary of classification algorithm implementation} 
\label{tab:table}
\end{table}


\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \toprule
Method & Abbreviation & Hyperparameters & R Package \\ 
  \midrule
Random Undesampling & RUS &   & ROSE \cite{rose} \\ 
  Random Oversampling & ROS &   & ROSE \cite{rose} \\ 
  Synthetic Majority Over Sampling & SMOTE & $k=5$ & IRIC \cite{iric} \\ 
  SMOTE - Edited Nearest Neighbours & SENN & $k_1=5$, $k_2 = 3$ & IRIC \cite{iric} \\ 
   \hline
{\footnotesize Note:}
 \\ {\footnotesize * $k$: the number of nearest neighbors in implementation of SMOTE.} \\ {\footnotesize * $k_1$: the number of nearest neighbors in the SMOTE step of SMOTE-ENN.} \\ {\footnotesize * $k_2$: the number of nearest neighbors in the ENN step of SMOTE-ENN}\end{tabular}
\caption{Summary of imbalance corrections implementation} 
\label{tab:table}
\end{table}
